The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-17 script/en/chapters/cryptomethods.tex
CT script: Minor corrections, de + en; final now for Beta04

% --------------------------------------------------------------------------
\section[Symmetric encryption]
{Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} you can execute the following modern
    symmetric encryption algorithms 
    (using the menu path {\bf Crypt \textbackslash{} Symmetric (modern)}): \\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm).
  }
\nopagebreak
For {\em symmetric} encryption \index{Encryption!symmetric} sender and
recipient must be in possession of a common (secret) key which they have
exchanged before actually starting to communicate. The sender uses this
key to encrypt the message and the recipient uses it to decrypt it.
\par \vskip + 3pt

@@

Up-to-the-minute procedure is the symmetric AES standard. The associated
Rijndael algorithm was declared winner of the AES award on October 2nd, 2000
and thus succeeds the DES procedure.

@@

% --------------------------------------------------------------------------
\subsection{New results about cryptanalysis of AES}
\index{Cryptanalysis} \label{NeueAES-Analyse}

@@

% --------------------------------------------------------------------------
% \newpage
\section[Hybrid procedures]{Hybrid procedures\footnotemark}
\footnotetext{%
Within CrypTool\index{CrypTool} you can find this technique using the menu
path {\bf Crypt \textbackslash{} Hybrid}: 
There you can follow the single steps and its dependencies with concrete
numbers. The variant with RSA as the asymmetric algorithm is graphically
visualized; the variant with ECC uses the standard dialogs. In both cases
AES is used as the symmetric algorithm.
}
\label{CM_Hybrid-procedures}
\index{Hybrid procedure}

@@

Compared to public-key ciphers based on mathematics like RSA, the structure of AES,
and most other modern symmetric ciphers, is very complex and cannot be explained
as easily as RSA.

@@

How to use Mini-AES is exhaustively described at the Sage reference page\\
\url{http://www.sagemath.org/doc/reference/sage/crypto/block_cipher/miniaes.html}.

>>>>

% --------------------------------------------------------------------------
\section[Symmetric encryption]
{Symmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool\index{CrypTool} you can execute the following modern
    symmetric encryption algorithms 
    (using the menu path {\bf Crypt \textbackslash{} Symmetric (modern)}): \\
    IDEA, RC2, RC4, DES (ECB), DES (CBC), Triple-DES (ECB), Triple-DES (CBC),
    MARS (AES candidate), RC6 (AES candidate), Serpent (AES candidate), 
    Twofish (AES candidate), Rijndael (official AES algorithm).
    \index{AES}
  }
\nopagebreak
For {\em symmetric} encryption \index{Encryption!symmetric} sender and
recipient must be in possession of a common (secret) key which they have
exchanged before actually starting to communicate. The sender uses this
key to encrypt the message and the recipient uses it to decrypt it.
\par \vskip + 3pt

@@

Up-to-the-minute procedure is the symmetric AES standard\index{AES}. The associated
Rijndael algorithm was declared winner of the AES award on October 2nd, 2000
and thus succeeds the DES procedure.\index{AES}

@@

% --------------------------------------------------------------------------
\subsection{New results about cryptanalysis of AES}
\label{NeueAES-Analyse}
\index{Cryptanalysis}
\index{AES}

@@

% --------------------------------------------------------------------------
% \newpage
\section[Hybrid procedures]{Hybrid procedures\footnotemark}
\footnotetext{%
Within CrypTool\index{CrypTool} you can find this technique using the menu
path {\bf Crypt \textbackslash{} Hybrid}: 
There you can follow the single steps and its dependencies with concrete
numbers. The variant with RSA as the asymmetric algorithm is graphically
visualized; the variant with ECC uses the standard dialogs. In both cases
AES is used as the symmetric algorithm.\index{AES}
}
\label{CM_Hybrid-procedures}
\index{Hybrid procedure}

@@

Compared to public-key ciphers based on mathematics like RSA, the structure of AES,
and most other modern symmetric ciphers, is very complex and cannot be explained
as easily as RSA.\index{AES}

@@

How to use Mini-AES is exhaustively described at the Sage reference page:\\
\url{http://www.sagemath.org/doc/reference/sage/crypto/block_cipher/miniaes.html}.

### 2009-08-17 script/en/chapters/digitalsignatures.tex
CT script: Minor corrections, de + en; final now for Beta04

\vspace{12pt}
The aim of digital signatures is to guarantee the following two points:
\begin{itemize}
 \item User authenticity: \index{Authenticity!user} \\
      It can be checked whether a message really does
come from a particular person.
 \item Message integrity:  \index{Message integrity} \\
      It can be checked whether the message has been
changed (on route).
\end{itemize}

@@

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. Because a signature is about
as long as the straight data stream to be signed, the procedure we just described
has in practice, however, a decisive disadvantage. The signature is
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{Performance}, we apply
a cryptographic hash function\footnote{%
Hash functions\index{Hash function} are implemented within 
CrypTool\index{CrypTool} at several places.\\
Using menus {\bf Individual Procedures \textbackslash{} Hash} and
            {\bf Analysis \textbackslash{} Hash}
you have the possibilities
% hier die items nicht einruecken!
\begin{list}{\textbullet}{\leftmargin10pt\addtolength{\itemsep}{-1.0\baselineskip}}
%\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hash value,\\
\item to calculate a key from a password according to the PKCS\#5 
      standard\index{PKCS\#5}, \\
\item to calculate HMACs from a text and a secret key, and\\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions\index{Collision}.
\end{list}
} to the document -- before signing. The output of the hash function will 
then be signed.

% --------------------------------------------------------------------------
\vskip + 15pt
\vskip + 30pt
%\newpage
\begin{center}
\fbox{\parbox{15cm}{{\em Stanislaw Lem\index{Lem, Stanislaw}\footnotemark:}\newline 
We can make everything out of this world, but we cannot create a world,
where humans in some ten thousand years can think: 'Ok, now it is enough.
Everything should stay like it is. Let's do no changes any more, don't do
inventions any more, because it cannot become better, and if, then we
don't want this.'
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{This was the answer of Stanislaw Lem
to heavy critics at his philosophical main book ~``Summa Technologiae'',
1964, where he thought about the possibility of an evolution creating
artificial intelligence.\\}

>>>>

\vspace{12pt}
\noindent The aim of digital signatures is to guarantee the following two points:
\begin{itemize}
 \item User authenticity: \index{Authenticity!user} \\
      It can be checked whether a message really does
come from a particular person.
 \item Message integrity:  \index{Message integrity} \\
      It can be checked whether the message has been
changed (on route).
\end{itemize}

@@

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. The procedure we just described
has in practice, however, a decisive disadvantage. The signature would be
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{Performance}, we apply
a cryptographic hash function\footnote{%
Hash functions\index{Hash function} are implemented within 
CrypTool\index{CrypTool} at several places.\\
Using menus {\bf Individual Procedures \textbackslash{} Hash} and
            {\bf Analysis \textbackslash{} Hash}
you have the possibilities
% hier die items nicht einruecken!
\begin{list}{\textbullet}{\leftmargin10pt\addtolength{\itemsep}{-1.0\baselineskip}}
%\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hash value,\\
\item to calculate a key from a password according to the PKCS\#5 
      standard\index{PKCS\#5}, \\
\item to calculate HMACs from a text and a secret key, and\\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions\index{Collision}.
\end{list}
} to the document -- before signing. The output of the hash function will 
then be signed.

% --------------------------------------------------------------------------
\vskip + 45pt
\begin{center}
\fbox{\parbox{15cm}{{\em Stanislaw Lem\index{Lem, Stanislaw}\footnotemark:}\newline 
We can make everything out of this world, but we cannot create a world,
where humans in some ten thousand years can think: 'Ok, now it is enough.
Everything should stay like it is. Let's do no changes any more, don't do
inventions any more, because it cannot become better, and if, then we
don't want this.'
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{This was the answer of Stanislaw Lem
to heavy critics at his philosophical main book ~``Summa Technologiae'',
1964, where he thought about the possibility of an evolution creating
artificial intelligence.\\}

### 2009-08-17 script/en/chapters/ellipticcurves.tex
CT script: Minor corrections, de + en; final now for Beta04

CrypTool\index{CrypTool} offers elliptic curves for the digital signature function%
\footnote{%
The dialog box, which appears in CrypTool\index{CrypTool} after clicking the menu
{\bf Digital Signatures/PKI \textbackslash{} Sign Message},
offers the EC methods ECSP-DSA and ECSP-NR.
} and for ECC-AES hybrid encryption%
\footnote{%
Within CrypTool\index{CrypTool} you can find this technique using the menu
path {\bf Crypt \textbackslash{} Hybrid}.
}.

It implements the basic algorithms for group operations, for generating elliptic
curves, for importing and exporting parameters for elliptic curves over finite
fields with $p$ ($p$ prime) elements. The algorithms have been implemented in
ANSI C and comply with draft no. 8 of the IEEE P1363 work group {\em Standard
Specifications for Public Key Cryptography}

@@

\url{http://www.sagemath.org/doc/constructions/elliptic_curves.html}%
\footnote{%
Sage samples can be found at:\\
- the Published Worksheets at \url{http://www.sagenb.org/pub/} like\\
- Elliptic Curve\\
  \url{http://www.sagenb.org/home/pub/606/} \\
- Elliptic Curve El Gamal\\
  \url{http://www.sagenb.org/home/pub/104/} \\
- or at the "Elliptic Curve Cryptography (ECC) Tutorial"\\
  \url{http://www.williamstein.org/simuw06/notes/notes/node12.html}
}.

There is an exhaustive ECC tutorial by Maike Massierer,
called \hyperlink{ec:Web-Link:Sage_Massierer}{ECC script}.
This interactive introduction to Elliptic Curve Cryptography is build up as
a Sage notebook%
\footnote{%
If you run your own Sage environment, you have to compile (e.g. with Sage 4.1)
the package kash3-2008-07-31.spkg, because the ECC notebook of Mrs. Massierer
needs it.
}.

This notebook\footnote{%
Starting the notebook in a browser: Enter "notebook()" at the Sage commandline.\\
- In your browser you normally first have to login.\\
- If you click in the above right corner at "Published", you will see our 8 worksheets.\\
You should then generate you own local (under Linux in your home area) copy by clicking
a worksheet (then click "Edit a copy" in the left upper corner).\\
- Sometimes it's necessary at the beginning to click in the left upper corner
"Action -> Evaluate all".\\
- Sometimes the window contains many (blue) error messages. This normally can be
solved by clicking the small "\%hide" string: Then you get the code behind the
graphics. Simply generate the graphics again with Shift-Enter.
},\footnote{%
  Since 2008 this tutorial can be found at
    \url{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}.\\
  - To logon in Siegen you have to allow port 8000 and Cookies.\\
  - The fonts from jsMAah are used. They should be installed into your browser for a better layout.
    After installing these fonts you can see the jsMath symbol at the lower border
    of your browser. If you click this symbol you can find the download page for the TTIF fonts.
    We are sorry, that this has to be done at every PC where his notebook runs in a browser.
}
consists of 8 chapters (title page plus 7 chapters)
% "Titelseite" mit Inhaltsverzeichnis plus 7 Kapitel
and aims to let even beginners understand what elliptic curves are:

@@

\begin{enumerate}
   \item Interactive introduction to elliptic curves and elliptic curve cryptography
         with Sage\index{Sage} by Maike Massierer and the CrypTool team,\\
         \href{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}
      {\texttt{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}} \\
          ECC Tutorial as Sage Notebook \\
         Version 1.01, October 2008
       \hypertarget{ec:Web-Link:Sage_Massierer}{}

   \item Certicom Online Tutorial, \index{Certicom},\\
         \href{http://www.certicom.com/resources/ecc_tutorial/ecc_tutorial.html}
      {\texttt{http://www.certicom.com/resources/ecc\_tutorial/ecc\_tutorial.html}}

>>>>

CrypTool offers elliptic curves for the digital signature function%
\footnote{%
The dialog box, which appears in CrypTool\index{CrypTool} after clicking the menu
{\bf Digital Signatures/PKI \textbackslash{} Sign Message},
offers the EC methods ECSP-DSA and ECSP-NR.
} and for ECC-AES hybrid encryption%
\footnote{%
Within CrypTool\index{CrypTool} you can find this technique using the menu
path {\bf Crypt \textbackslash{} Hybrid}.
}.

It implements the basic algorithms for group operations, for generating elliptic
curves, for importing and exporting parameters for elliptic curves over finite
fields with $p$ elements ($p$ prime). The algorithms have been implemented in
ANSI C and comply with draft no. 8 of the IEEE P1363 work group {\em Standard
Specifications for Public Key Cryptography}

@@

\url{http://www.sagemath.org/doc/constructions/elliptic_curves.html}%
\footnote{%
Sage samples can be found at:\\
- the Published Worksheets at \url{http://www.sagenb.org/pub/} like\\
- Elliptic Curve: \url{http://www.sagenb.org/home/pub/606/} \\
- Elliptic Curve El Gamal: \url{http://www.sagenb.org/home/pub/104/} \\
- or at the "Elliptic Curve Cryptography (ECC) Tutorial"\\
  \url{http://www.williamstein.org/simuw06/notes/notes/node12.html}
}.

There is an exhaustive ECC tutorial by Maike Massierer,
called \hyperlink{ec:Web-Link:Sage_Massierer}{ECC script}.
This interactive introduction to Elliptic Curve Cryptography is build up as
a Sage notebook%
\footnote{%
If you run your own Sage environment, you have to compile
the package ``kash3-2008-07-31.spkg'' (e.g. with Sage 4.1),
because the ECC notebook of Mrs. Massierer needs it.
}.

This notebook\footnote{%
Starting the Sage notebook in a browser: Enter ``notebook()'' at the Sage commandline.\\
- In your browser you normally first have to login.\\
- If you click in the above right corner at ``Published'', you will see our 8 worksheets.\\
You should then generate your own local copy (under Linux in your home area) by clicking
a worksheet (then click ``Edit a copy'' in the left upper corner).\\
- Sometimes it's necessary at the beginning to click in the left upper corner
  \verb#Action -> Evaluate all#.\\
- Sometimes the window contains many (blue) error messages. This normally can be
solved by clicking the ``\%hide'' string: Then you get the code behind the
graphics. Simply generate the graphics again with Shift-Enter.
},\footnote{%
  Since 2008 this tutorial can be found at
    \url{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}.\\
  - To logon in Siegen you have to allow port 8000 and Cookies.\\
  - This tutorial uses the fonts from jsMath. They should be installed into your browser
    for a better layout.
    See \url{http://www.math.union.edu/~dpvc/jsMath/}
    and \url{http://pubpages.unh.edu/~jsh3/jsMath/}.\\
    After installing these fonts you can see the jsMath symbol at the lower border
    of your browser. If you click this symbol you can find the download page for the TIFF fonts.
    We are sorry, that this has to be done at every PC where his notebook runs in a browser.
}
consists of 8 chapters (title page plus 7 chapters)
% "Titelseite" mit Inhaltsverzeichnis plus 7 Kapitel
and aims to let even beginners understand what elliptic curves are:

@@

\begin{enumerate}
        \hypertarget{ec:Web-Link:Sage_Massierer}{} 
   \item Interactive introduction to elliptic curves and elliptic curve cryptography
        with Sage\index{Sage} by Maike Massierer and the CrypTool team,\\
        \href{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}
        {\texttt{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}} \\
        ECC Tutorial as Sage Notebook \\
        Version 1.01, October 2008
   \item Certicom Online Tutorial, \index{Certicom},\\
        \href{http://www.certicom.com/resources/ecc_tutorial/ecc_tutorial.html}
        {\texttt{http://www.certicom.com/resources/ecc\_tutorial/ecc\_tutorial.html}}

### 2009-08-17 script/en/chapters/numbertheory.tex
CT script: Sage parts finished in de + en; several corrections
CT script: Minor corrections, de + en; final now for Beta04

% --------------------------------------------------------------------------
\subsection{Complexity}\label{complexity}\index{Complexity}

@@

% --------------------------------------------------------------------------
\vskip +20pt
\hypertarget{RSA-200-chap3}{}
\paragraph{RSA-200} \label{RSA-200} \index{RSA-200}\mbox{}
%\hypertarget{RSA-200-chap3}{}

@@

% --------------------------------------------------------------------------
\subsection{The Diffie-Hellman key exchange protocol}
\index{Diffie, Whitfield} 
\index{Hellman, Martin} 
\index{Key agreement (key exchange)!Diffie-Hellman}
\index{Diffie-Hellman}

@@

\vskip +10 pt
\input{figures/DH-en.latex}
\vskip +10 pt

{\bf Procedure:}

\noindent Alice and Bob want to negotiate a secret session key $S$ via a channel that may be intercepted. 
\begin{itemize}
\item[{\bf 1.}] They select a prime number $p$ and a random number $g$ and exchange this information openly.
\item[{\bf 2.}] Alice now selects $a$, a random number less than $p$ and keeps it secret.

@@

  \item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                  {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                  of $e$ modulo $J(n)$. We can then ``throw away'' $J(n)$.
    \begin{itemize}
        \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the private key $S$ (only $d$ must be kept secret).
    \end{itemize}

@@

The default way to crack a given RSA ciphertext $C$ would be to use the public key
of the recipient and to try to factorize $n$. Then you can go through the steps 2 and 3
and calculate the private key $e$, which is normally used to decrypt a ciphertext.

@@

\noindent If you don't want to use factorization but ask the question like in classic
encryption, you may want to find out:
How many possible private keys $(n, d)$ are there for a given key size range
$n \in \{a, b\}$? %% (this means, only a part of each public key $(n, e)$ is known).

>>>>

% --------------------------------------------------------------------------
\subsection{Complexity}
\label{complexity}
\index{Complexity}

@@

% --------------------------------------------------------------------------
\vskip +20pt
\hypertarget{RSA-200-chap3}{}
\paragraph{RSA-200} \label{RSA-200} \index{RSA-200}\mbox{}

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection{The Diffie-Hellman key exchange protocol}
\index{Diffie, Whitfield} 
\index{Hellman, Martin} 
\index{Key agreement (key exchange)!Diffie-Hellman}
\index{Diffie-Hellman}

@@

\vskip +10 pt
\input{figures/DH-en.latex}
\vskip +20 pt

\noindent {\bf Procedure:}\par  %\par bewirkt Zeilenumbruch
\nopagebreak
\noindent Alice and Bob want to negotiate a secret session key $S$ via
a channel that may be intercepted. 
\begin{itemize}
\item[{\bf 1.}] They select a prime number $p$ and a random number $g$ and exchange this information openly.
\item[{\bf 2.}] Alice now selects $a$, a random number less than $p$ and keeps it secret.

@@

  \item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                  {\rm ~(mod~} J(n))$,\\
		  i.e. $d$ is the multiplicative inverse of $e$ modulo $J(n)$.
		  We can then ``throw away'' $J(n)$.
    \begin{itemize}
        \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the private key $S$ (only $d$ must be kept secret).
    \end{itemize}

  \item[{\bf 4.}] For encryption, the message represented as a (binary) number
                  is divided into parts such that each part of 
                  the number represents a number less than $n$.

@@

The default way to crack a given RSA ciphertext $C$ would be to use the public key
of the recipient and to try to factorize $n$. Then you can go through the steps 2 and 3
and generate the private key $e$, which is normally used to decrypt a ciphertext.

@@

\noindent If you don't want to use factorization but ask the question like in classic
encryption, you may want to find out:
How many possible private keys $(n, d)$ are there for a given key size range
$n \in [a, b]$? %% (this means, only a part of each public key $(n, e)$ is known).
